---
title: "Validation of logistic regression risk prediction models"
always_allow_html: true
output:
  github_document:
    toc: true
    toc_depth: 5
  keep_text: true
  pandoc_args: --webtex
---
  
## Steps
  
The steps taken in this file are:   
1. To develop a logistic regression risk prediction model.  
2. To assess the performance of the model in terms of calibration, discrimination and overall prediction error. We calculate the apparent, internal (optimism-corrected) validation and the external validation.  
3. To assess the potential clinical utility the model using decision curve analysis.  

### Installing and loading packages and import data

The following libraries are used in this file, the code chunk below will a) check whether you already have them installed, b) install them for you if not already present, and c) load the packages into the session.

```{r setup, include=FALSE}
# Knitr options
knitr::opts_chunk$set(
  fig.retina = 3,
  fig.path = "imgs/",
  echo = FALSE
)
```

```{r, wdlib, message=FALSE, warning=FALSE, echo=TRUE}
# Use pacman to check whether packages are installed, if not load
if (!require("pacman")) install.packages("pacman")
library(pacman)

pacman::p_load(
  rio,
  rms,
  riskRegression,
  plotrix,
  knitr,
  splines,
  kableExtra,
  gtsummary,
  boot,
  tidyverse,
  rsample,
  gridExtra,
  webshot
)

# Import data ------------------
rdata <- readRDS(here::here("Data/rdata.rds"))
vdata <- readRDS(here::here("Data/vdata.rds"))

```

### Data description
Men with metastatic non-seminomatous testicular cancer can often be cured nowadays by cisplatin based chemotherapy. After chemotherapy, surgical resection is a generally accepted treatment to remove remnants of the initial metastases, since residual tumor may still be present. In the absence of tumor, resection has no therapeutic benefits, while it is associated with hospital admission, and risks of permanent morbidity and mortality. Logistic regression models were developed to predict the presence of residual tumor, combining  well-known predictors, such as the histology of the primary tumor, pre-chemotherapy levels of tumor markers, and (reduction in) residual mass size.  
We first consider a data set (rdata) with 544 patients to develop a prediction model that includes 5 predictors. We then extend this model with the pre-chemotherapy level of the tumor marker lactate dehydrogenase (LDH). This illustrates ways to assess the incremental
value of a marker. LDH values were log transformed, after standardizing by dividing by the local upper levels of normal values, after examination of non-linearity with restricted cubic spline functions.  


We  first consider a data set with 544 patients to develop a prediction model that includes 5 predictors (rdata).  In a later study, we externally validated the 5 predictor model in 273 patients from a tertiary referral center (vdata). We illustrate ways to assess the usefulness of a model in a new setting.  
We then extend the developed model with the pre-chemotherapy level of the tumor marker lactate dehydrogenase (LDH). Since the validation data (vdata) did not have information about LDH, we assess the prediction performances of the basic model in the development (rdata) and in the validation data (vdata). 

Thus, we loaded the development data (rdata) and the validation data (vdata).  

More details about development and validation data are provided in the manuscript ["Assessing the performance of prediction models: a framework for some traditional and novel measures"](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3575184/) by Steyerberg et al. (2010).

#### Descriptive statistics

```{r, import}
rsel <- rdata |> select(-patkey)
vsel <- vdata |> select(-patkey)
rsel$dt <- 1
vsel$dt <- 2
cdata <- rbind.data.frame(rsel, vsel)
cdata$dt <- factor(
  cdata$dt,
  levels = c(1, 2),
  labels = c("Development data", "Validation data")
)
cdata$tum_res <- factor(cdata$tum_res,
                        levels = c(0, 1),
                        labels = c("No", "Yes"))
label(cdata$tum_res) <- "Residual tumor at postchemotherapy resection"
label(cdata$ter_pos) <- "Residual tumor at postchemotherapy resection "
label(cdata$preafp) <- "Prechemotherapy AFP elevated?"
label(cdata$prehcg) <- "Prechemotherapy HCG elevated?"
label(cdata$sqpost) <- "Square root of postchemotherapy mass size"
label(cdata$reduc10) <- "Reduction in mass size per 10%"
label(cdata$lnldhst) <- "Prechemotherapy LDH"
units(cdata$sqpost) <- "mm"
units(cdata$lnldhst) <- "log(LDH/upper limit of local normal value)"
```


```{r tab1_bis, warning=FALSE, message=FALSE}
gtsummary::tbl_summary(
  data = cdata ,
  label = list(
    sqpost ~ "Square root of postchemotherapy mass size (mm)", 
    reduc10 ~ "Reduction in mass size per 10%",
    lnldhst ~ "Prechemotherapy LDH"),
  by = "dt",
  type = all_continuous() ~ "continuous2",
  statistic = all_continuous() ~ c(
     "{mean} ({sd})",
     "{median} ({min}, {max})"
 )
) |>
  gtsummary::as_kable_extra() |>
  kableExtra::kable_styling("striped")

```

## Goal 1 - develop a logistic regression risk prediction model

### 1.1 Check non-linearity of continuous predictors

Here we investigate the potential non-linear relation between continuous predictors (i.e. age and size) and the outcomes. We apply three-knot restricted cubic splines using `rms::rcs()` function (details are given in e.g. Frank Harrell's book 'Regression Model Strategies (second edition)', page 27. We assess the potential non-linearity graphically (plotting the two continuous predictors against the log odds (XB or linear predictor) of both event types. Also, we compare the models with and without splines based on the AIC.

<details>
  <summary>Click to expand code</summary>
```{r,ff, warning=FALSE, fig.align='center', eval=FALSE, echo=TRUE}

# Models without splines
dd <- datadist(rdata, adjto.cat = "first")
options(datadist = "dd")
fit_lrm <- lrm(tum_res ~ 
               ter_pos + preafp + prehcg + 
               sqpost + reduc10,
               data = rdata, x = T, y = T)
options(datadist = NULL)

# Models with splines
dd <- datadist(rdata, adjto.cat = "first")
options(datadist = "dd")
fit_lrm_rcs <- lrm(tum_res ~ 
                   ter_pos + preafp + prehcg + 
                   rcs(sqpost, 3) + rcs(reduc10, 3) ,
                   data = rdata, x = T, y = T)
options(datadist = NULL)


# print(fit_lrm_rcs)
# print(summary(fit_lrm_rcs))
# print(anova(fit_lrm_rcs))
P_lrm_sqpost_rcs <- Predict(fit_lrm_rcs, "sqpost")
P_lrm_reduc10_rcs <- Predict(fit_lrm_rcs, "reduc10")
options(datadist = NULL)

oldpar <- par(mfrow = c(1, 2), mar = c(5, 5, 1, 1))
par(xaxs = "i", yaxs = "i", las = 1)
# Square root of postchemotherapy mass size
plot(
  P_lrm_sqpost_rcs$sqpost,
  P_lrm_sqpost_rcs$yhat,
  type = "l",
  lwd = 2,
  col = "blue",
  bty = "n",
  xlab = "Square root of postchemotherapy mass size",
  ylab = "log Odds",
  ylim = c(-3, 1.5),
  xlim = c(0, 12)
)
polygon(
  c(P_lrm_sqpost_rcs$sqpost, rev(P_lrm_sqpost_rcs$sqpost)),
  c(P_lrm_sqpost_rcs$upper, rev(P_lrm_sqpost_rcs$lower)),
  col = "grey75",
  border = FALSE
)
par(new = TRUE)
plot(
  P_lrm_sqpost_rcs$sqpost,
  P_lrm_sqpost_rcs$yhat,
  type = "l",
  lwd = 2,
  col = "blue",
  bty = "n",
  xlab = "Square root of postchemotherapy mass size",
  ylab = "log Odds",
  ylim = c(-3, 1.5),
  xlim = c(0, 12)
)

# Reduction in mass size per 10%
plot(
  P_lrm_reduc10_rcs$reduc10,
  P_lrm_reduc10_rcs$yhat,
  type = "l",
  lwd = 2,
  col = "blue",
  bty = "n",
  xlab = "Reduction in mass size per 10%",
  ylab = "log Odds",
  xlim = c(-5, 10),
  ylim = c(-2, 3)
)
polygon(
  c(P_lrm_reduc10_rcs$reduc10, rev(P_lrm_reduc10_rcs$reduc10)),
  c(P_lrm_reduc10_rcs$upper, rev(P_lrm_reduc10_rcs$lower)),
  col = "grey75",
  border = FALSE
)
par(new = TRUE)
plot(
  P_lrm_reduc10_rcs$reduc10,
  P_lrm_reduc10_rcs$yhat,
  type = "l",
  lwd = 2,
  col = "blue",
  bty = "n",
  xlab = "Reduction in mass size per 10%",
  ylab = "log Odds",
  xlim = c(-5, 10),
  ylim = c(-2, 3)
)
options(datadist = NULL)
par(oldpar)
```
</details>

```{r,ff, warning=FALSE, fig.align='center', eval=TRUE}
```

```{r, res_aic, fig.align='center'}
res_AIC <- matrix(c(
  AIC(fit_lrm), AIC(fit_lrm_rcs)
),
byrow = T,
ncol = 2,
nrow = 1,
dimnames =
  list(
    c(
      "Residual tumor at postchemotherapy resection"
    ),
    c(
      "AIC without splines",
      "AIC with splines"
    )
  )
)
kable(res_AIC, row.names = TRUE) |>
  kable_styling("striped", position = "center")
```

Both the graphical comparison and the AIC comparison suggested no relevant departure from linear relations between the continuous predictors (square root of post-chemotherapy mass size and reduction in mass size) and the risk of residual tumor at post-chemotherapy resection.  


### 1.2 Examine the fit of the models

+ Logistic regression risk prediction model without LDH

```{r, summary_lrm1, fig.align='center',warning=FALSE}
dd <- datadist(rdata)
options(datadist = "dd")
options(prType = "html")
fit_lrm1 <- lrm(tum_res ~ 
               ter_pos + preafp + prehcg + 
               sqpost + reduc10,
               data = rdata, x = T, y = T)
print(fit_lrm1)
# print(summary(fit_lrm))
options(datadist = NULL)
```

+ Logistic regression risk prediction model with LDH

```{r, summary_lrm2, fig.align='center',warning=FALSE}
dd <- datadist(rdata)
options(datadist = "dd")
options(prType = "html")
fit_lrm2 <- lrm(tum_res ~ 
               ter_pos + preafp + prehcg + 
               sqpost + reduc10 + lnldhst,
               data = rdata, x = T, y = T)
print(fit_lrm2)
# print(summary(fit_lrm2))
options(datadist = NULL)
```

The coefficients of the models indicated that positive tumor teratoma, elevated prechemoterapy AFP levels, elevated prechemoterapy HCG levels, postchemotherapy mass size (mm) (expressed in square root) are associated with higher risk to residual tumor after resection. Reduction in mass size is associated with a reduced risk to have residual tumor after resection. 

### 1.5 Plot of predictors vs estimated in the validation data

To get further insight into the effect of the covariates, we plot the covariate values observed in the validation set against the estimated absolute risk of having residual tumor after resection. This gives an idea of the size of the effects.

<details>
  <summary>Click to expand code</summary>
```{r, plot_risk, fig.align='center',warning=FALSE, echo=TRUE, eval=FALSE}
# Models -------------
dd <- datadist(rdata)
options(datadist = "dd")
options(prType = "html")
fit_lrm1 <- lrm(tum_res ~ 
               ter_pos + preafp + prehcg + 
               sqpost + reduc10,
               data = rdata, x = T, y = T)
# print(summary(fit_lrm))
options(datadist = NULL)


# Calculate predicted probabilities in the validation data
vdata$pred <- predict(fit_lrm1,
              type = "fitted.ind",
              newdata = vdata)

# Positive teratoma tumor
oldpar <- par(mfrow = c(2, 3))
par(xaxs = "i", yaxs = "i", las = 1)
plot(vdata$ter_pos,
  vdata$pred,
  bty = "n",
  ylim = c(0, 1.2),
  xlab = "Positive teratoma tumor",
  ylab = "Estimated risk"
)
# Elevated AFP
par(xaxs = "i", yaxs = "i", las = 1)
plot(vdata$preafp,
  vdata$pred,
  bty = "n",
  ylim = c(0, 1.2),
  xlab = "Elevated AFP levels",
  ylab = "Estimated risk",
  cex.lab = .85
)

# Elevated HCG
par(xaxs = "i", yaxs = "i", las = 1)
plot(vdata$prehcg,
  vdata$pred,
  bty = "n",
  ylim = c(0, 1.2),
  xlab = "Elevated HGC levels",
  ylab = "Estimated risk",
  cex.lab = .85
)

# Postchemotherapy mass size
par(xaxs = "i", yaxs = "i", las = 1)
plot(vdata$sqpost,
  vdata$pred,
  bty = "n",
  xlim = c(0, 20),
  ylim = c(0, 1.2),
  xlab = "Square root of postchemotherapy mass size",
  ylab = "Estimated risk",
  cex.lab = .85
)
lines(
  lowess(vdata$sqpost, vdata$pred),
  col = "red",
  lwd = 2
)

# Reduction mass size
par(xaxs = "i", yaxs = "i", las = 1)
plot(vdata$reduc10,
  vdata$pred,
  bty = "n",
  xlim = c(0, 10),
  ylim = c(0, 1.2),
  xlab = "Reduction mass size",
  ylab = "Estimated risk",
  cex.lab = .85
)
lines(
  lowess(vdata$reduc10, vdata$pred),
  col = "red",
  lwd = 2
)

```
</details>

```{r, plot_risk, fig.align='center',warning=FALSE, echo=FALSE, eval=TRUE}
```

## Goal 2 - Assessing performance of a logistic regression risk prediction model

Here we evaluate the performance of the prediction model in terms of discrimination, calibration and overall prediction error. 
We assess the prediction performance of the developed model not including LDH internally and in an external data.   


### 2.1 Discrimination

We here calculate:  

+ The c-statistic: it is a rank order statistic for predictions against true outcomes. The concordance (c) statistic is the most commonly used performance measure to indicate the discriminative ability of generalized linear regression models. For a binary outcome, c is identical to the area under the Receiver Operating Characteristic (ROC) curve, which plots the sensitivity (true positive rate) against 1 – (false positive rate) for consecutive cutoffs for the probability of an outcome. Accurate predictions discriminate between those with and those without the outcome.

+ Discrimination slope: it can be used as a simple measure for how well subjects with and without the outcome are separated. It is calculated as the absolute difference in average predictions for those with and without the outcome. Visualization is readily possible with a box plot or a histogram, which will show less overlap between those with and those without the outcome for a better discriminating model. 
 

More details are in ["Assessing the performance of prediction models: a framework for some traditional and novel measures"](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3575184/) by Steyerberg et al. (2010);


<details>
  <summary>Click to expand code</summary>
```{r, discrimination, warning=FALSE, message=FALSE, echo=TRUE}

# Models
dd <- datadist(rdata, adjto.cat = "first")
options(datadist = "dd")
fit_lrm <- lrm(tum_res ~ 
               ter_pos + preafp + prehcg + 
               sqpost + reduc10,
               data = rdata, x = T, y = T)
options(datadist = NULL)


# C-index
# Apparent validation
val_rdata <- rcorr.cens(predict(fit_lrm), 
                        S = rdata$tum_res)

# Load function to calculate confidence intervals of c-statistic
source(here::here('Functions/c_stat_ci.R'))
c_dev <- c_stat_ci(val_rdata)


# Internal validation (bootstrapping)
val_opt <- validate(fit = fit_lrm, B = 500)
c_optimism <- round(0.5 * 
                      (1 +
                      val_opt["Dxy","index.corrected"]), 2)

# External validation
val_vdata <- rcorr.cens(
  predict(fit_lrm, newdata = vdata), 
  S = vdata$tum_res)

c_val <- c_stat_ci(val_vdata)

res_c <- rbind(c_dev, c_val)

# Discrimination slope
# Load function to calculate discrimination slope
source(here::here('Functions/discr_slope.R'))
dslope_rdata <- discr_slope(fit_lrm,
                            y = rdata$tum_res,
                            new_data = rdata)
dslope_vdata <- discr_slope(fit_lrm,
                            y = vdata$tum_res,
                            new_data = vdata)

# Bootstrap confidence intervals for discrimination slope
rboot <- bootstraps(rdata, B = 1000)
vboot <- bootstraps(vdata, B = 1000)

discr_slope_boot <- function(split) {
  discr_slope(fit = fit_lrm,
              y = analysis(split)$tum_res,
              new_data = analysis(split))
}

rboot <- rboot |>
  mutate(dslope = map_dbl(splits, discr_slope_boot))

vboot <- vboot |>
  mutate(dslope = map_dbl(splits, discr_slope_boot))

# Optimism-corrected discrimination slope
source(here::here("Functions/internal_cv_lrm.R"))

optim_measures <- bootstrap_cv_lrm(
  db = rdata, 
  B = 50, # set B lower to speed up the computation
  outcome = "tum_res", 
  formula = "tum_res ~ 
                  ter_pos + preafp + prehcg + 
                  sqpost + reduc10", 
  formula_score = "tum_res ~ 1")

```
</details>
```{r, discrimination, fig.align='center', warning=FALSE, eval=TRUE}
```

```{r, res_disc, fig.align='center',echo=FALSE}
alpha <- .05
k <- 2
res_discr <- matrix(c(
  c_dev,
  
  c_optimism,
  NA,
  NA,
  
  c_val,
  
  dslope_rdata,
  quantile(rboot$dslope, probs = alpha / 2),
  quantile(rboot$dslope, probs = 1 - alpha / 2),
  
  optim_measures["Discrimination slope corrected"],
  NA,
  NA,
  
  dslope_vdata,
  quantile(vboot$dslope, probs = alpha / 2),
  quantile(vboot$dslope, probs = 1 - alpha /2)
  ),
 
  nrow = 2,
  ncol = 9,
  byrow = T,
  
  dimnames = list(
    c("C-statistic", "Discrimination slope"),
    rep(c("Estimate", "Lower .95 ", "Upper .95"), 3)
  )
)
res_discr <- round(res_discr, 2) # Digits
kable(res_discr) |>
  kable_styling("striped", position = "center") |>
  add_header_above(c(" " = 1, 
                     "Apparent" = 3, 
                     "Internal" = 3, 
                     "External" = 3))
```
C-statistic was 0.82 (95% confidence interval, CI: 0.78-0.85), and 0.78 (95% CI: 0.73-0.84) in the development and validation data, respectively. Internal cross-validation based on optimism-corrected bootstrapping showed a C-statistic of 0.81.  

Discrimination slope was 0.30 (95% CI: 0.27-0.32), and 0.24 (95% CI: 0.17-0.29) for the development and validation data, respectively.

### 2.2 Calibration
Calibration refers to the agreement between observed outcomes and predictions. For example, if we predict a 20% risk of residual tumor for a testicular cancer patient, the observed frequency of tumor should be approximately 20 out of 100 patients with such a prediction.

Different level of calibration can be estimated: mean, weak, and moderate calibration according to the calibration hierarchy defined by Van Calster et al. [here](https://www.sciencedirect.com/science/article/pii/S0895435615005818).  

#### 2.2.1 Mean calibration
Mean calibration refers how systematically the model might under- or over- predicts the actual risk.   

The mean calibration can be estimated:  

+ using the Observed and Expected ratio. The observed number of events is the sum of the events (or cases) present in the data. The expected is estimated summing the predicted probability of the event estimated by the model. Ratio equals to 1 indicates perfect (mean) calibration, values lower or greater than 1 indicate over- and under- prediction, respectively.

+ calibration intercept (or calibration-in-the-large): indicates the extent that predictions are systematically too low or too high.  


<details>
  <summary>Click to expand code</summary>
```{r, mean_cal, fig.align='center',echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
# Models
dd <- datadist(rdata, adjto.cat = "first")
options(datadist = "dd")
fit_lrm <- lrm(tum_res ~ 
               ter_pos + preafp + prehcg + 
               sqpost + reduc10,
               data = rdata, x = T, y = T)
options(datadist = NULL)

# Calibration-in-the-large
lp <- predict(fit_lrm, newdata = vdata)
vdata$y <- as.numeric(vdata$tum_res) - 1 # convert outcome to numeric
cal_intercept <- glm(y  ~ offset(lp), 
                     family = binomial,
                     data = vdata)
intercept_CI <- confint(cal_intercept) # confidence intervals

# Observed/Expected ratio
Obs <- sum(vdata$y)
vdata$pred <- predict(fit_lrm, 
                      newdata = vdata,
                      type = "fitted.ind")
Expct <- sum(vdata$pred)

OE <- Obs / Expct

```
</details>

```{r, mean_cal, fig.align='center', warning=FALSE, eval=TRUE}
```

```{r, res_mean_cal, fig.align='center',echo=FALSE}
alpha <- .05
k <- 2
res_mean_cal <- matrix(c(
    cal_intercept$coefficients,
    
    intercept_CI[1],
    intercept_CI[2],
    
    OE,

    OE*exp(-qnorm(1 - alpha / 2) * sqrt(1 / Obs)),
    OE*exp(qnorm(1 - alpha / 2) * sqrt(1 / Obs))
),
  nrow = 2,
  ncol = 3, 
  byrow = T,
  dimnames = list(
    c("Calibration intercept", "O/E ratio"),
    c("Estimate", "Lower .95 ", "Upper .95"))
)

res_mean_cal <- round(res_mean_cal, 2) # Digits
kable(res_mean_cal) |>
  kable_styling("striped", position = "center") 

```
Both calibration intercept and O/E ratio showed good mean calibration.
The prediction model did not systematically over or underestimate the actual risk. 

#### 2.2.2 Weak calibration
The term ‘weak’ refers to the limited flexibility in assessing calibration. We are essentially summarizing calibration of the observed proportions of outcomes versus predicted probabilities using only two parameters i.e. a straight line. In other words, perfect weak
calibration is defined as mean calibration ratio and calibration slope of unity(or calibration intercept equals to zero). 
The calibration slope indicates the overall strength of the linear predictor (LP), which can be interpreted as the level of overfitting
(slope <1) or underfitting (slope>1). A value of slope smaller than 1 can also be interpreted as reflecting a need for shrinkage of regression coefficients in a prediction model. 

<details>
  <summary>Click to expand code</summary>
```{r, weak_cal, fig.align='center',echo=TRUE, eval=FALSE ,message=FALSE, warning=FALSE}
# Models
dd <- datadist(rdata, adjto.cat = "first")
options(datadist = "dd")
fit_lrm <- lrm(tum_res ~ 
               ter_pos + preafp + prehcg + 
               sqpost + reduc10,
               data = rdata, x = T, y = T)
options(datadist = NULL)

# Calibration slope
vdata$lp <- predict(fit_lrm, newdata = vdata)
cal_slope <- glm(y  ~ lp,
                 family = binomial,
                 data = vdata)
slope_CI <- confint(cal_slope) # Confidence interval
```
</details>

```{r, weak_cal, fig.align='center', warning=FALSE, eval=TRUE}
```

```{r, res_weak_cal, fig.align='center',echo=FALSE}
k <- 2
res_weak_cal <- cbind(
  "Estimate" = cal_slope$coefficients[2],
  "Lower .95" = slope_CI[1],
  "Upper .95" = slope_CI[2]
)
rownames(res_weak_cal) <- "Calibration slope"

res_weak_cal <- round(res_weak_cal, 2) # Digits
kable(res_weak_cal) |>
  kable_styling("striped", position = "center") 
```

#### 2.2.3 Moderate calibration
Moderate calibration concerns whether among patients with the same predicted risk, the observed event rate equals the predicted risk. A graphical assessment of calibration is possible with predictions on the x-axis, and the outcome on the y-axis. Perfect predictions should be on the 45° line.  For binary outcomes, the plot contains only 0 and 1 values for the y-axis. Smoothing techniques can be used to estimate
the observed probabilities of the outcome (p(y=1)) in relation to the predicted probabilities, e.g. using the lowess algorithm. The observed probabilities can be also estimated using a secondary logistic regression model using the predicted probability as predictor modelled with splines. We may however expect that the specific type of smoothing may affect the graphical impression, especially in smaller data sets.

<details>
  <summary>Click to expand code</summary>
```{r, moder_cal, fig.align='center',echo=TRUE, eval=FALSE, warning=FALSE, message=FALSE}
# Models
dd <- datadist(rdata, adjto.cat = "first")
options(datadist = "dd")
fit_lrm <- lrm(tum_res ~ 
               ter_pos + preafp + prehcg + 
               sqpost + reduc10,
               data = rdata, x = T, y = T)
options(datadist = NULL)

## Calibration plot
vdata$y <- as.numeric(vdata$tum_res) - 1
vdata$pred <- predict(fit_lrm,
                      newdata = vdata,
                      type = "fitted.ind")

# Calibration based on a secondary logistic regression
fit_cal <- glm(y ~ pred,
               x = T,
               y = T,
               data = vdata)
# NOTE: we can also model using splines (e.g. y ~ rcs(pred, 5),
# restricted cubic splies with five knots

cal_obs <- predict(fit_cal,  
                   type = "response",
                   se.fit = TRUE)
alpha <- .05
dt_cal <- cbind.data.frame("obs" = cal_obs$fit,
                           
                           "lower" = 
                             cal_obs$fit - 
                             qnorm(1 - alpha / 2)*cal_obs$se.fit,
                           
                           "upper" = cal_obs$fit + 
                             qnorm(1 - alpha / 2)*cal_obs$se.fit,
                           
                           "pred" = vdata$pred)
dt_cal <- dt_cal[order(dt_cal$pred),]

cal_lowess <- lowess(vdata$pred, vdata$y, iter = 0)
par(xaxs = "i", yaxs = "i", las = 1)
plot(cal_lowess,
     type = "l",
     xlim = c(0, 1),
     ylim = c(0, 1),
     xlab = "Predicted probability",
     ylab = "Actual probability",
     bty = "n",
     lwd = 2,
     main = "Calibration plot")
lines(dt_cal$pred, dt_cal$obs, lwd = 2, lty = 2)
lines(dt_cal$pred, dt_cal$lower, lwd = 2, lty = 3)
lines(dt_cal$pred, dt_cal$upper, lwd = 2, lty = 3)
abline(a = 0, b = 1, col = "gray")
legend("bottomright",
       c("Ideal", "Non - parametricLowess", 
         "Logistic",
         "95% confidence interval"),
       lwd = c(1, 2, 2, 2),
       lty = c(1, 1, 2, 3),
       col = c("gray", "black", "black", "black"),
       bty = "n")

# Calibration measures ICI, E50, E90 based on secondary logistic regression
res_calmeas <-
  c(
    "ICI" = mean(abs(dt_cal$obs - dt_cal$pred)),
    "E50" = median(abs(dt_cal$obs - dt_cal$pred)),
    "E90" = unname(quantile(abs(dt_cal$obs - dt_cal$pred), 
                            probs = .90))
)

## Bootstrap confidence intervals
## for the calibration  measures (ICI, E50, E90) ------
alpha <- .05
B <- 1000 # Set B = 2000 although it takes more time
set.seed(2022)
vboot <- bootstraps(vdata, times = B)

# Bootstrap calibration measures
numsum_boot <- function(split) {
  
  pred <- predict(fit_lrm,
                  type = "fitted.ind",
                  newdata = analysis(split))

 
# Estimate actual risk - basic model
  vcal <- glm(y ~ pred,
              x = T,
              y = T,
              data = analysis(split)
) 
 
cal_obs_boot <- predict(vcal, 
                        type = "response",
                        se.fit = TRUE,
                        newdata = analysis(split))
 # Save objects needed
db_cal_boot <- data.frame(
  "obs" = cal_obs_boot$fit,
  
  "lower" = cal_obs_boot$fit - 
    qnorm(1 - alpha / 2)*cal_obs_boot$se.fit,
                           
  "upper" = cal_obs_boot$fit +  
    qnorm(1 - alpha / 2)*cal_obs_boot$se.fit,
                      
  "pred" = pred
)

absdiff_boot <- abs(db_cal_boot$obs - db_cal_boot$pred)

res_cal_boot <- data.frame(
  "ICI" = mean(absdiff_boot),
  "E50" = quantile(absdiff_boot, probs = .5),
  "E90" = quantile(absdiff_boot, probs = .9)
)
}
numsum_b <- vboot |>
  mutate(num_cal_boot = map(splits, numsum_boot),
         
         ICI = map_dbl(num_cal_boot, ~ .x$ICI),
         
         E50 = map_dbl(num_cal_boot, ~ .x$E50),
         
         E90 = map_dbl(num_cal_boot, ~ .x$E90)
         )
```
</details>
```{r, moder_cal, fig.align='center', warning=FALSE, eval=TRUE}
```

```{r, res_numcal, message=FALSE,warning=FALSE, fig.align='center',include=TRUE, echo=FALSE}
alpha <- .05
res_numcal <- matrix(c(res_calmeas["ICI"],
                       quantile(numsum_b$ICI, probs = alpha / 2),
                       quantile(numsum_b$ICI, 
                                probs = 1 - alpha / 2),
                       
                       res_calmeas["E50"],
                       quantile(numsum_b$E50, probs = alpha / 2),
                       quantile(numsum_b$E50, 
                                probs = 1 - alpha / 2),
                       
                       res_calmeas["E90"],
                       quantile(numsum_b$E90, probs = alpha / 2),
                       quantile(numsum_b$E90, 
                                probs = 1 - alpha / 2)
                       ),
                     nrow = 1,
                     ncol = 9,
                     byrow = T,
                     dimnames = list(
                       c("External data"),
                       rep(
                         c("Estimate", "Lower.95", "Upper.95"),
                         3))
)
res_numcal <- round(res_numcal, 2)
kable(res_numcal) |>
  kable_styling("striped", position = "center") |>
  add_header_above(c(" " = 1,  
                     "ICI" = 3, 
                     "E50" = 3,
                     "E90" = 3))
```
Calibration measures (i.e., ICI, E50, E90) using a 'secondary' logistic regression to estimate the observed probability of the event indicate good calibration.  
Calibration measures might also be calculated using the non-parametric method (e.g., lowess) to estimate the observed probability of the event. In this setting, the user can use the results of ```stats::lowess()``` function to calculate the absolute difference between the estimated observed and predicted probabilities. Another nice function to provide the calibration plot and the corresponding performance measures is ```rms::val.prob()```. Another alternative is ```riskRegression::Score()```.  


### 2.3 Overall performance measures
The overall performance measures generally estimate the distance between the predicted outcome and actual outcome.  
We calculate the Brier Score, and the scaled Brier scale (also indicated as index of prediction accuracy) and the corresponding confidence intervals.

Some confidence intervals are calculated using the bootstrap percentile method.

<details>
  <summary>Click to expand code</summary>
```{r, overall, warning=FALSE, message=FALSE, eval=FALSE, echo=TRUE}

# Models -------------------
dd <- datadist(rdata, adjto.cat = "first")
options(datadist = "dd")
fit_lrm <- lrm(tum_res ~ 
               ter_pos + preafp + prehcg + 
               sqpost + reduc10,
               data = rdata, x = T, y = T)
options(datadist = NULL)


# Overall performance measures ----------------

# Development data
score_rdata <- Score(
  list("Development set" = fit_lrm),
  formula = tum_res ~ 1,
  data = rdata,
  conf.int = TRUE,
  metrics = c("auc", "brier"),
  summary = c("ipa"),
  plots = "calibration"
)

# Validation data
score_vdata <- Score(
  list("Validation set" = fit_lrm),
  formula = tum_res ~ 1,
  data = vdata,
  conf.int = TRUE,
  metrics = c("auc", "brier"),
  summary = c("ipa"),
  plots = "calibration"
)

# Optimism-corrected Brier and scaled Brier (IPA)
source(here::here("Functions/internal_cv_lrm.R"))

optim_measures <- bootstrap_cv_lrm(
  db = rdata, 
  B = 50, # set B lower to speed up the computation
  outcome = "tum_res", 
  formula = "tum_res ~ 
                  ter_pos + preafp + prehcg + 
                  sqpost + reduc10", 
  formula_score = "tum_res ~ 1")


# Bootstrap confidence intervals for scaled Brier score -------
# For Brier, bootstrap should be computationally faster when
# data has more than 2000 rows (see ?riskRegression::Score).
# Our data has 1000 row so we will need only bootstrap to calculate
# confidence intervals of the scaled Brier (IPA) since
# it is not provided by riskRegression::Score() function.

# Bootstrapping data
set.seed(2022)
B <- 50 # number of bootstrap samples
rboot <- bootstraps(rdata, times = B)
vboot <- bootstraps(vdata, times = B)

# Score functions in any bootstrap data
score_boot <- function(split) {
  Score(
    list("Logistic" = fit_lrm),
    formula = tum_res ~ 1,
    data = analysis(split),
    conf.int = TRUE,
    metrics = c("auc","brier"),
    summary = c("ipa"),
    plots = "calibration"
  )
}

# Development data
rboot <- rboot |> mutate(
  score = map(splits, score_boot),
  scaled_brier = map_dbl(score, function(x) {
    x$Brier$score[model == "Logistic"]$IPA
  })
)
# Validation data
vboot <- vboot |> mutate(
  score = map(splits, score_boot),
  scaled_brier = map_dbl(score, function(x) {
    x$Brier$score[model == "Logistic"]$IPA
  })
)

```
</details>

```{r, overall, fig.align='center', warning=FALSE, eval=TRUE}
```

```{r, res_ov, fig.align='center',echo=FALSE}
# Table overall measures
alpha <- .05
k <- 2
res_ov <- matrix(c(
                 score_rdata$Brier$score$Brier[2],
                 score_rdata$Brier$score$lower[2],
                 score_rdata$Brier$score$upper[2],
                 
                 optim_measures["Brier corrected"],
                 NA,
                 NA,
                 
                 score_vdata$Brier$score$Brier[2],
                 score_vdata$Brier$score$lower[2],
                 score_vdata$Brier$score$upper[2],
                 
                 score_rdata$Brier$score$IPA[2],
                 quantile(rboot$scaled_brier, probs = alpha / 2),
                 quantile(rboot$scaled_brier, probs = 1 - alpha / 2),
                 
                 optim_measures["IPA corrected"],
                 NA,
                 NA,
                 
                 score_vdata$Brier$score$IPA[2],
                 quantile(vboot$scaled_brier, probs = alpha / 2),
                 quantile(vboot$scaled_brier, probs = 1 - alpha / 2)
                 ),
                 nrow = 2,
                 ncol = 9,
                 byrow = T,
                 dimnames = list(
                   c("Brier", "scaled Brier (IPA)"),
                   rep(c("Estimate", "Lower.95", "Upper.95"), 3))
)

res_ov <- round(res_ov, 2) # Digits
kable(res_ov) |>
  kable_styling("striped", position = "center") |>
  add_header_above(c(" " = 1, 
                     "Apparent" = 3, 
                     "Internal" = 3, 
                     "External" = 3))

```


## Goal 3 -  Clinical utility
Discrimination and calibration measures are essential to assess the prediction performance but insufficient to evaluate the potential clinical utility of a risk prediction model for decision making. When new markers are available, clinical utility assessment evaluates whether the extended model helps to improve decision making.  
Clinical utility is measured by the net benefit that includes the number of true positives and the number of false positives.
Generally, in medicine, clinicians accepts to treat a certain number of patients for which interventions are unnecessary to be event free for a given time horizon. So, false negatives (the harm of not being event free for a given time horizon) are more important than false positives (the harm of unnecessary interventions). Thus, net benefit is the number of true positives classifications minus the false positives classifications weighted by a factor related to the harm of not preventing the event versus unnecessary interventions. The weighting is derived from the threshold probability to the event of interest (e.g. residual tumor). For example, a threshold of 20% implies that additional interventions for 4 patients of whom one would have experience the event if untreated is acceptable (thus treating 9 unnecessary patients). This strategy is compared with the strategies of treat all and treat none patients. If overtreatment is harmful, a higher threshold should be used. 

The net benefit is calculated as:  
  
<img src="https://render.githubusercontent.com/render/math?math=%5Chuge%7B%5Cfrac%7BTP%7D%7Bn%7D-%5Cfrac%7BFP%7D%7Bn%7D*%5Cfrac%7Bp_t%7D%7B1-p_t%7D%7D">
  
*TP*=true positive patients   
*FP*=false positive patients  
*n*=number of patients and *p*<sub>t</sub> is the risk threshold.  

  
The decision curve is calculated as follows:
  
1. Choose a time horizon;
2. Specify a risk threshold which reflects the ratio between harms and benefit of an additional intervention;
3. Calculate the number of true positive and false positive given the threshold specified in (2);
4. Calculate the net benefit of the risk prediction model;
5. Plot net benefit on the *y-axis* against the risk threshold on the *x-axis*;
6. Repeat steps 2-4 for each model consideration;
7. Repeat steps 2-4 for the strategy of assuming all patients are treated;
8. Draw a straight line parallel to the *x-axis* at y=0 representing the net benefit associated with the strategy of assuming that all patients are not treated.

Given some thresholds, the model/strategy with higher net benefit represents the one that potentially improves  clinical decision making. However, poor discrimination and calibration lead to lower net benefit.

More details are available in the paper of Vickers et al. [here](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2577036/).  

<details>
  <summary>Click to expand code</summary>
```{r, dca, message=FALSE,warning=FALSE, fig.align='center',echo=TRUE,eval=FALSE}
# Run the stdca function to calculate the net benefit and the elements needed to develop decision curve analysis
source(here::here("Functions/dca.R"))

# Models ------------------------------
dd <- datadist(rdata, adjto.cat = "first")
options(datadist = "dd")
fit_lrm <- lrm(tum_res ~ 
               ter_pos + preafp + prehcg + 
               sqpost + reduc10,
               data = rdata, x = T, y = T)
options(datadist = NULL)

# Development data
# calculation estimated risk
rdata$y <- as.numeric(rdata$tum_res) - 1
rdata$pred <- predict(fit_lrm,
                      type = "fitted.ind")
rdata <- as.data.frame(rdata)
dca_rdata <- dca(
  data = rdata,
  outcome = "y",
  predictors = "pred",
  graph = FALSE
)
# Decision curves plot
oldpar <- par(
  xaxs = "i",
  yaxs = "i",
  las = 1,
  mar = c(6.1, 5.8, 4.1, 2.1),
  mgp = c(4.25, 1, 0)
)
plot(dca_rdata$net.benefit$threshold,
  dca_rdata$net.benefit$pred,
  type = "l",
  lwd = 2,
  lty = 1,
  xlab = "",
  ylab = "Net Benefit",
  xlim = c(0, 1),
  ylim = c(-0.10, 0.8),
  bty = "n",
  xaxt = "n"
)
legend("topright",
  c("Treat all", "Treat none", "Prediction model"),
  lwd = c(2, 2, 2),
  lty = c(1, 2, 1),
  col = c("darkgray", "black", "black"),
  bty = "n",
  cex = .85
)
lines(dca_rdata$net.benefit$threshold,
  dca_rdata$net.benefit$none,
  type = "l",
  lwd = 2,
  lty = 4
)
lines(dca_rdata$net.benefit$threshold,
  dca_rdata$net.benefit$all,
  type = "l",
  lwd = 2,
  col = "darkgray"
)
axis(1,
  at = c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1.0)
)
# axis(1,
#   pos = -0.145,
#   at = c(0.1, 0.2, 0.3, 0.4, 0.5),
#   labels = c("1:9", "1:4", "3:7", "2:3", "1:1")
# )
mtext("Threshold probability", 1, line = 2)
# mtext("Harm to benefit ratio", 1, line = 5)
title("Development data")
par(oldpar)


# Validation data
# Predicted probability calculation
vdata$y <- as.numeric(vdata$tum_res) - 1
vdata$pred <- predict(fit_lrm,
                       newdata = vdata,
                       type = "fitted.ind")
vdata <- as.data.frame(vdata)
# Run decision curve analysis
# Development data
# Model without PGR
dca_vdata <- dca(
  data = vdata,
  outcome = "y",
  predictors = "pred",
  graph = FALSE,
)
# Decision curves plot
oldpar <- par(
  xaxs = "i",
  yaxs = "i",
  las = 1,
  mar = c(6.1, 5.8, 4.1, 2.1),
  mgp = c(4.25, 1, 0)
)
plot(dca_vdata$net.benefit$threshold,
  dca_vdata$net.benefit$pred,
  type = "l",
  lwd = 2,
  lty = 1,
  xlab = "",
  ylab = "Net Benefit",
  xlim = c(0, 1),
  ylim = c(-0.10, 0.8),
  bty = "n",
  xaxt = "n"
)
lines(dca_vdata$net.benefit$threshold,
  dca_vdata$net.benefit$none,
  type = "l",
  lwd = 2,
  lty = 4
)
lines(dca_vdata$net.benefit$threshold,
  dca_vdata$net.benefit$all,
  type = "l",
  lwd = 2,
  col = "darkgray"
)
legend("topright",
  c("Treat all", "Treat none", "Prediction model"),
  lwd = c(2, 2, 2),
  lty = c(1, 2, 1),
  col = c("darkgray", "black", "black"),
  bty = "n",
  cex = .85
)
axis(1,
  at = c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1.0)
)
#axis(1,
#  pos = -0.145,
#  at = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1.0),
#  labels = c("1:9", "1:4", "3:7", "2:3", "1:1")
#)
mtext("Threshold probability", 1, line = 2)
# mtext("Harm to benefit ratio", 1, line = 5)
title("Validation data")
par(oldpar)
```
</details>

```{r, dca, message=FALSE,warning=FALSE, fig.align='center',echo=FALSE,eval=TRUE}
```

A cut-off of 20% implies a relative weight of 1:4 for false-positive decisions against true-positive decisions. In the development data, the Net Benefit (NB) was 0.439. This means that the model might identify approximately 44 patients out of 100 who may have residuals tumors and then tumor resection might be useful. If we would do resection in all, the NB would however be similar: 0.437 so the clinical utility of the model at 20% threshold is limited. 

Similar results were estimated during external validation. In the validation data, the NB of the risk regression model was
0.653. If we would do resection in all patients, the NB would be similar 0.652. 

However, the decision curve shows that the NB would be much larger for higher threshold values, i.e. patients accepting higher risks of
residual tumor. 

Moreover, potential net benefit can be defined in terms of reduction of avoidable interventions (e.g adjuvant chemotherapy per 100 patients) by:
  
<img src="https://render.githubusercontent.com/render/math?math=%5Chuge%7B%5Cfrac%7BNB_%7Bmodel%7D%20-%20NB_%7Ball%7D%7D%7B(p_t%2F%20(1-p_t))%7D*100%7D%0A">  
  
where *NB*<sub>model</sub> is the net benefit of the prediction model, *NB*<sub>all</sub> is the net benefit of the strategy treat all and $p_{t}$ is the risk threshold.

## Reproducibility ticket

```{r repro_ticket, echo=TRUE}
sessionInfo()
```
