---
title: "Validation of logistic regression risk prediction models"
always_allow_html: true
output:
  github_document:
    toc: true
    toc_depth: 5
  keep_text: true
  pandoc_args: --webtex
---
  
## Steps
  
The steps taken in this file are:   
1. To develop a logistic regression risk prediction model.
2. To assess the performance of the model in terms of calibration, discrimination and overall prediction error. We calculate the apparent, internal (optimism-corrected) validation and the external validation.
3. To assess the potential clinical utility the model using decision curve analysis.

### Installing and loading packages and import data

The following libraries are used in this file, the code chunk below will a) check whether you already have them installed, b) install them for you if not already present, and c) load the packages into the session.

```{r setup, include=FALSE}
# Knitr options
knitr::opts_chunk$set(
  fig.retina = 3,
  fig.path = "imgs/",
  echo = FALSE
)
```

```{r, wdlib, message=FALSE, warning=FALSE, echo=TRUE}
# Use pacman to check whether packages are installed, if not load
if (!require("pacman")) install.packages("pacman")
library(pacman)

pacman::p_load(
  rio,
  rms,
  pec,
  plotrix,
  knitr,
  splines,
  kableExtra,
  gtsummary,
  boot,
  tidyverse,
  rsample,
  gridExtra,
  webshot
)

# Import data ------------------
rdata <- readRDS(here::here("Data/rdata.rds"))
vdata <- readRDS(here::here("Data/vdata.rds"))

```

We loaded the development data (rdata) and the validation data (vdata).
More details about development and validation data are provided in the manuscript ["Assessing the performance of prediction models: a framework for some traditional and novel measures"](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3575184/) by Steyerberg et al. (2010).

### Descriptive statistics

```{r, import}
rsel <- rdata %>% select(-patkey)
vsel <- vdata %>% select(-patkey)
rsel$dt <- 1
vsel$dt <- 2
cdata <- rbind.data.frame(rsel, vsel)
cdata$dt <- factor(
  cdata$dt,
  levels = c(1, 2),
  labels = c("Development data", "Validation data")
)
cdata$tum_res <- factor(cdata$tum_res,
                        levels = c(0, 1),
                        labels = c("No", "Yes"))
label(cdata$tum_res) <- "Residual tumor at postchemotherapy resection"
label(cdata$ter_pos) <- "Residual tumor at postchemotherapy resection "
label(cdata$preafp) <- "Prechemotherapy AFP elevated?"
label(cdata$prehcg) <- "Prechemotherapy HCG elevated?"
label(cdata$sqpost) <- "Square root of postchemotherapy mass size"
label(cdata$reduc10) <- "Reduction in mass size per 10%"
label(cdata$lnldhst) <- "Prechemotherapy LDH"
units(cdata$sqpost) <- "mm"
units(cdata$lnldhst) <- "log(LDH/upper limit of local normal value)"
```


```{r tab1_bis, warning=FALSE, message=FALSE}
gtsummary::tbl_summary(
  data = cdata ,
  label = list(
    sqpost ~ "Square root of postchemotherapy mass size (mm)", 
    reduc10 ~ "Reduction in mass size per 10%",
    lnldhst ~ "Prechemotherapy LDH"),
  by = "dt",
  type = all_continuous() ~ "continuous2",
  statistic = all_continuous() ~ c(
     "{mean} ({sd})",
     "{median} ({min}, {max})"
 )
) %>%
  gtsummary::as_kable_extra() %>%
  kableExtra::kable_styling("striped")

```

## Goal 1 - develop a logistic regression risk prediction model

### 1.1 Check non-linearity of continuous predictors

Here we investigate the potential non-linear relation between continuous predictors (i.e. age and size) and the outcomes. We apply three-knot restricted cubic splines using `rms::rcs()` function (details are given in e.g. Frank Harrell's book 'Regression Model Strategies (second edition)', page 27. We assess the potential non-linearity graphically (plotting the two continuous predictors against the log relative hazards (XB or linear predictor) of both event types. Also, we compare the models with and without splines based on the AIC.

<details>
  <summary>Click to expand code</summary>
```{r,ff, warning=FALSE, fig.align='center', eval=FALSE, echo=TRUE}

# Models without splines
dd <- datadist(rdata, adjto.cat = "first")
options(datadist = "dd")
fit_lrm <- lrm(tum_res ~ 
               ter_pos + preafp + prehcg + 
               sqpost + reduc10,
               data = rdata, x = T, y = T)
options(datadist = NULL)

# Models with splines
dd <- datadist(rdata, adjto.cat = "first")
options(datadist = "dd")
fit_lrm_rcs <- lrm(tum_res ~ 
                   ter_pos + preafp + prehcg + 
                   rcs(sqpost, 3) + rcs(reduc10, 3) ,
                   data = rdata, x = T, y = T)
options(datadist = NULL)


# print(fit_lrm_rcs)
# print(summary(fit_lrm_rcs))
# print(anova(fit_lrm_rcs))
P_lrm_sqpost_rcs <- Predict(fit_lrm_rcs, "sqpost")
P_lrm_reduc10_rcs <- Predict(fit_lrm_rcs, "reduc10")
options(datadist = NULL)

oldpar <- par(mfrow = c(1, 2), mar = c(5, 5, 1, 1))
par(xaxs = "i", yaxs = "i", las = 1)
# Square root of postchemotherapy mass size
plot(
  P_lrm_sqpost_rcs$sqpost,
  P_lrm_sqpost_rcs$yhat,
  type = "l",
  lwd = 2,
  col = "blue",
  bty = "n",
  xlab = "Square root of postchemotherapy mass size",
  ylab = "log Odds",
  ylim = c(-3, 1.5),
  xlim = c(0, 12)
)
polygon(
  c(P_lrm_sqpost_rcs$sqpost, rev(P_lrm_sqpost_rcs$sqpost)),
  c(P_lrm_sqpost_rcs$upper, rev(P_lrm_sqpost_rcs$lower)),
  col = "grey75",
  border = FALSE
)
par(new = TRUE)
plot(
  P_lrm_sqpost_rcs$sqpost,
  P_lrm_sqpost_rcs$yhat,
  type = "l",
  lwd = 2,
  col = "blue",
  bty = "n",
  xlab = "Square root of postchemotherapy mass size",
  ylab = "log Odds",
  ylim = c(-3, 1.5),
  xlim = c(0, 12)
)

# Reduction in mass size per 10%
plot(
  P_lrm_reduc10_rcs$reduc10,
  P_lrm_reduc10_rcs$yhat,
  type = "l",
  lwd = 2,
  col = "blue",
  bty = "n",
  xlab = "Reduction in mass size per 10%",
  ylab = "log Odds",
  xlim = c(-5, 10),
  ylim = c(-2, 3)
)
polygon(
  c(P_lrm_reduc10_rcs$reduc10, rev(P_lrm_reduc10_rcs$reduc10)),
  c(P_lrm_reduc10_rcs$upper, rev(P_lrm_reduc10_rcs$lower)),
  col = "grey75",
  border = FALSE
)
par(new = TRUE)
plot(
  P_lrm_reduc10_rcs$reduc10,
  P_lrm_reduc10_rcs$yhat,
  type = "l",
  lwd = 2,
  col = "blue",
  bty = "n",
  xlab = "Reduction in mass size per 10%",
  ylab = "log Odds",
  xlim = c(-5, 10),
  ylim = c(-2, 3)
)
options(datadist = NULL)
par(oldpar)
```
</details>

```{r,ff, warning=FALSE, fig.align='center', eval=TRUE}
```

```{r, res_aic, fig.align='center'}
res_AIC <- matrix(c(
  AIC(fit_lrm), AIC(fit_lrm_rcs)
),
byrow = T,
ncol = 2,
nrow = 1,
dimnames =
  list(
    c(
      "Residual tumor at postchemotherapy resection"
    ),
    c(
      "AIC without splines",
      "AIC with splines"
    )
  )
)
kable(res_AIC, row.names = TRUE) %>%
  kable_styling("striped", position = "center")
```

Both the graphical comparison and the AIC comparison suggested no relevant departure from linear relations between the continuous predictors (square root of post-chemotherapy mass size and reduction in mass size) and the risk of residual tumor at post-chemotherapy resection.  


### 1.2 Examine the fit of the models

+ Logistic regression risk prediction model without LDH

```{r, summary_lrm1, fig.align='center',warning=FALSE}
dd <- datadist(rdata)
options(datadist = "dd")
options(prType = "html")
fit_lrm1 <- lrm(tum_res ~ 
               ter_pos + preafp + prehcg + 
               sqpost + reduc10,
               data = rdata, x = T, y = T)
print(fit_lrm1)
# print(summary(fit_lrm))
options(datadist = NULL)
```

+ Logistic regression risk prediction model with LDH

```{r, summary_lrm2, fig.align='center',warning=FALSE}
dd <- datadist(rdata)
options(datadist = "dd")
options(prType = "html")
fit_lrm2 <- lrm(tum_res ~ 
               ter_pos + preafp + prehcg + 
               sqpost + reduc10 + lnldhst,
               data = rdata, x = T, y = T)
print(fit_lrm2)
# print(summary(fit_lrm2))
options(datadist = NULL)
```

The coefficients of the models indicated that positive tumor teratoma, elevated prechemoterapy AFP levels, elevated prechemoterapy HCG levels, postchemotherapy mass size (mm) (expressed in square root) are associated with higher risk to residual tumor after resection. Reduction in mass size is associated with a reduced risk to have residual tumor after resection. 

### 1.5 Plot of predictors vs estimated  in the validation data

To get further insight into the effect of the covariates, we plot the covariate values observed in the validation set against the estimated absolute risk of having residual tumor after resection. This gives an idea of the size of the effects.

<details>
  <summary>Click to expand code</summary>
```{r, plot_risk, fig.align='center',warning=FALSE, echo=TRUE, eval=FALSE}
# Models -------------
dd <- datadist(rdata)
options(datadist = "dd")
options(prType = "html")
fit_lrm1 <- lrm(tum_res ~ 
               ter_pos + preafp + prehcg + 
               sqpost + reduc10,
               data = rdata, x = T, y = T)
# print(summary(fit_lrm))
options(datadist = NULL)


# Calculate predicted probabilities in the validation data
vdata$pred <- predict(fit_lrm1,
              type = "fitted.ind",
              newdata = vdata)

# Positive teratoma tumor
oldpar <- par(mfrow = c(2, 3))
par(xaxs = "i", yaxs = "i", las = 1)
plot(vdata$ter_pos,
  vdata$pred,
  bty = "n",
  ylim = c(0, 1),
  xlab = "Positive teratoma tumor",
  ylab = "Estimated risk"
)
# Elevated AFP
par(xaxs = "i", yaxs = "i", las = 1)
plot(vdata$preafp,
  vdata$pred,
  bty = "n",
  ylim = c(0, 1),
  xlab = "Elevated AFP levels",
  ylab = "Estimated risk"
)

# Elevated HCG
par(xaxs = "i", yaxs = "i", las = 1)
plot(vdata$prehcg,
  vdata$pred,
  bty = "n",
  ylim = c(0, 1),
  xlab = "Elevated HGC levels",
  ylab = "Estimated risk"
)

# Postchemotherapy mass size
par(xaxs = "i", yaxs = "i", las = 1)
plot(vdata$sqpost,
  vdata$pred,
  bty = "n",
  xlim = c(0, 20),
  ylim = c(0, 1),
  xlab = "Square root of postchemotherapy mass size",
  ylab = "Estimated risk"
)
lines(
  lowess(vdata$sqpost, vdata$pred),
  col = "red",
  lwd = 2
)

# Reduction mass size
par(xaxs = "i", yaxs = "i", las = 1)
plot(vdata$reduc10,
  vdata$pred,
  bty = "n",
  xlim = c(0, 10),
  ylim = c(0, 1),
  xlab = "Reduction mass size",
  ylab = "Estimated risk"
)
lines(
  lowess(vdata$reduc10, vdata$pred),
  col = "red",
  lwd = 2
)

```
</details>

```{r, plot_risk, fig.align='center',warning=FALSE, echo=FALSE, eval=TRUE}
```

## Goal 2 - Assessing performance of a logistic regression risk prediction model

Here we evaluate the performance of the prediction model in terms of calibration, discrimination and overall prediction error. 


### 2.1 Calibration

We assess calibration by:

- The calibration plot as a graphical representation of calibration using
 the pseudo observations and the subdistribution hazard approach
 - Numerical summaries of calibration:  
 - The observed vs expected ratio (O/E ratio) ;  
 - The squared bias, i.e., the average squared difference between actual risks and risk predictions;
 - The integrated Calibration Index (ICI), i.e., the average absolute difference between actual risks and risk predictions;  
 - E50, E90 and Emax denote the median, 90th percentile and the maximum of the absolute differences between actual risks and risk predictions;  
 - Calibration intercept/slope estimated using pseudo observations:  
 - If on average the risk estimates equal the actual risks, the calibration intercept will be zero. A negative calibration intercept indicates that the risk estimates are on average too high and a positive intercept indicates that the risk estimates are on average too low.   
 - A calibration slope between 0 and 1 indicates overfitting of the model, i.e., too extreme predictions, both on the low and on the high         end. A calibration slope >1 indicates predictions do not show enough variation.  
      
#### 2.1.1 Calibration using pseudo observations

We calculate calibration plot and numerical summaries of calibration as
ICI, E50, E90, Emax and root squared bias using pseudo value approach.

##### 2.1.1.1 Calibration plot using pseudo observations


Calibration plot suggests that the prediction model seems to overestimate the actual risk, especially at the lower and higher values of the estimated risk.


#### 2.1.2 Calibration using the subdistribution hazard approach
Here we assess calibration using the subdistribution hazard approach (Austin et al.)

##### 2.1.2.1 Calibration plot using the subdistribution hazard approach

<details>
  <summary>Click to expand code</summary>
```{r, plot_sd, fig.align='center',warning=FALSE,message=FALSE,eval=FALSE,echo=TRUE}
# Models ----------
fit_csh <- CSC(Hist(time, status_num) ~
age + size +
  ncat + hr_status,
data = rdata,
fitter = "cph"
)


# useful objects
primary_event <- 1 # Set to 2 if cause 2 was of interest
horizon <- 5 # Set time horizon for prediction (here 5 years)


# Calibration plot (flexible regression approach) -------------------------

# Add estimated risk and complementary log-log of it to dataset
vdata$pred <- predictRisk(fit_csh,
  cause = primary_event,
  newdata = vdata,
  times = horizon
)
vdata$cll_pred <- log(-log(1 - pred))

# 5 knots seems to give somewhat equivalent graph to pseudo method with bw = 0.05
n_internal_knots <- 5 # Austin et al. advise to use between 3 (more smoothing, less flexible) and 5 (less smoothing, more flexible)
rcs_vdata <- ns(vdata$cll_pred, df = n_internal_knots + 1)
colnames(rcs_vdata) <- paste0("basisf_", colnames(rcs_vdata))
vdata_bis <- cbind.data.frame(vdata, rcs_vdata)

# Use subdistribution hazards (Fine-Gray) model
form_fgr <- reformulate(
  termlabels = colnames(rcs_vdata),
  response = "Hist(time, status_num)"
)

# Regress subdistribution of event of interest on cloglog of estimated risks
calib_fgr <- FGR(
  formula = form_fgr,
  cause = primary_event,
  data = vdata_bis
)

# Add observed and predicted together in a data frame
dat_fgr <- cbind.data.frame(
  "obs" = predict(calib_fgr, times = horizon, newdata = vdata_bis),
  "pred" = vdata$pred
)

# Calibration plot
dat_fgr <- dat_fgr[order(dat_fgr$pred), ]
par(xaxs = "i", yaxs = "i", las = 1)
plot(
  x = dat_fgr$pred,
  y = dat_fgr$obs,
  type = "l",
  xlim = c(0, 0.6),
  ylim = c(0, 0.6),
  xlab = "Predictions",
  ylab = "Estimated actual risk",
  bty = "n"
)
abline(a = 0, b = 1, lty = "dashed", col = "red")
title("Calibration plot using subdistribution hazard approach")
```
</details>

```{r, plot_sd, fig.align='center',warning=FALSE,message=FALSE,echo=FALSE,eval=TRUE}
```

Calibration plot suggests that the prediction model seems to overestimate the actual risk, especially at the lower and higher values of the estimated risk.

##### 2.1.2.2 Numerical summaries of calibration using the subdistribution hazard approach
Here we assess calibration using the subdistribution hazard approach (Austin et al.)

```{r, num_cal_sd, fig.align='center',warning=FALSE,message=FALSE}
# Numerical summary measures
diff_fgr <- dat_fgr$pred - dat_fgr$obs

numsum_fgr <- c(
  "ICI" = mean(abs(diff_fgr)),
  setNames(quantile(abs(diff_fgr), c(0.5, 0.9)), c("E50", "E90")),
  "Emax" = max(abs(diff_fgr)),
  "Root squared bias" = sqrt(mean(diff_fgr^2))
)

```

```{r, res_cal_sd, fig.align='center',warning=FALSE,message=FALSE,echo=FALSE}
k <- 4
res_num <- rbind(
  "Calibration measures - subdistribution " = numsum_fgr
)

res_num <- round(res_num, k)
kable(res_num) %>%
  kable_styling("striped", position = "center")
```
Numerical calibration measures identified overestimation of the risk especially in the higher values of the estimated actual risk.



#### 2.1.3 Observed and Expected ratio

<details>
  <summary>Click to expand code</summary>
```{r, oe_ratio, fig.align='center',warning=FALSE,message=FALSE,echo=TRUE}

# Models ----------
fit_csh <- CSC(Hist(time, status_num) ~
age + size +
  ncat + hr_status,
data = rdata,
fitter = "cph"
)


# useful objects
primary_event <- 1 # Set to 2 if cause 2 was of interest
horizon <- 5 # Set time horizon for prediction (here 5 years)

# Add estimated risk and complementary log-log of it to dataset
pred <- predictRisk(fit_csh,
  cause = primary_event,
  newdata = vdata,
  times = horizon
)

## Observed/Expected ratio --------------------------------------------
# First calculate Aalen-Johansen estimate (as 'observed')
obj <- summary(survfit(Surv(time, status) ~ 1,
  data = vdata
),
times = horizon
)

aj <- list(
  "obs" = obj$pstate[, primary_event + 1],
  "se" = obj$std.err[, primary_event + 1]
)


# Calculate O/E
OE <- aj$obs / mean(pred)

# For the confidence interval we use method proposed in Debray et al. (2017) doi:10.1136/bmj.i6460
k <- 2
alpha <- 0.05
OE_summary <- cbind(
  "OE" = OE,
  "Lower .95" = exp(log(OE - qnorm(1 - alpha / 2) * aj$se / aj$obs)),
  "Upper .95" = exp(log(OE + qnorm(1 - alpha / 2) * aj$se / aj$obs))
)

OE_summary <- round(OE_summary, k)
```
</details>

```{r, res_OE, echo=FALSE}
# OE, rootmeansquared, E50, E90, Emax, int, slope,

kable(OE_summary) %>%
  kable_styling("striped", position = "center")
```

Observed and expected ratio shown slight overestimation of the risk predicted by the model.

#### 2.1.4 Calibration intercept and slope using pseudo observations


<details>
  <summary>Click to expand code</summary>
```{r, cal_intslope_pseudo, fig.align='center',warning=FALSE,message=FALSE, echo=TRUE}

# Models ----------
fit_csh <- CSC(Hist(time, status_num) ~
age + size +
  ncat + hr_status,
data = rdata,
fitter = "cph"
)


# useful objects
primary_event <- 1 # Set to 2 if cause 2 was of interest
horizon <- 5 # Set time horizon for prediction (here 5 years)

# Predicted risk estimation
pred <- predictRisk(fit_csh,
  cause = primary_event,
  times = horizon,
  newdata = vdata
)


# Calibration plot (pseudo-obs approach) ----------------------------------
# First compute riskRegression::Score()
score_vdata <- Score(
  list("csh_validation" = fit_csh),
  formula = Hist(time, status_num) ~ 1,
  cens.model = "km",
  data = vdata,
  conf.int = TRUE,
  times = horizon,
  #  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)


## Calibration intercept and slope --------------------------------------
# Use pseudo-observations calculated by Score() (can alternatively use pseudo::pseudoci)
pseudos <- data.frame(score_vdata$Calibration$plotframe)

# Note:
# - 'pseudos' is the data.frame with ACTUAL pseudo-observations, not the smoothed ones
# - Column ID is not the id in vdata; it is just a number assigned to each row of
# the original validation data sorted by time and event indicator
pseudos$cll_pred <- log(-log(1 - pseudos$risk)) # add the cloglog risk ests

# Fit model for calibration intercept
fit_cal_int <- geese(
  pseudovalue ~ offset(cll_pred),
  data = pseudos,
  id = ID,
  scale.fix = TRUE,
  family = gaussian,
  mean.link = "cloglog",
  corstr = "independence",
  jack = TRUE
)

# Fit model for calibration slope
fit_cal_slope <- geese(
  pseudovalue ~ offset(cll_pred) + cll_pred,
  data = pseudos,
  id = ID,
  scale.fix = TRUE,
  family = gaussian,
  mean.link = "cloglog",
  corstr = "independence",
  jack = TRUE
)

# Perform joint test on intercept and slope
betas <- fit_cal_slope$beta
vcov_mat <- fit_cal_slope$vbeta
wald <- drop(betas %*% solve(vcov_mat) %*% betas)
# pchisq(wald, df = 2, lower.tail = FALSE)
```
</details>

```{r, res_cal_intslope, echo=FALSE}
k <- 2
res_cal <- rbind(

  # Value, confidence interval and test for calibration intercept
  "Intercept" = with(
    summary(fit_cal_int)$mean,
    c(
      "estimate" = estimate,
      `2.5 %` = estimate - qnorm(0.975) * san.se,
      `97.5 %` = estimate + qnorm(0.975) * san.se
    )
  ),

  # Value, confidence interval and test for calibration slope
  "Slope" = with(
    summary(fit_cal_slope)$mean["cll_pred", ],
    c(
      "estimate" = 1 + estimate,
      `2.5 %` = 1 + (estimate - qnorm(0.975) * san.se),
      `97.5 %` = 1 + (estimate + qnorm(0.975) * san.se)
    )
  )
)

res_cal <- round(res_cal, k)
kable(res_cal) %>%
  kable_styling("striped", position = "center")
```

The calibration intercept was estimated at -0.15 [95% CI -0.36 to 0.05] also pointing towards slight overestimation (though not statistically significant). This number for example means that for an estimated risk of 30%, the expected actual risk is 1-0.7^(exp(-0.15))=26%. The calibration slope was 1.22 [95% CI 0.84 to 1.60], which would indicate too homogeneous predictions but the wide confidence interval precludes any firm conclusions from it. The p-value for the joint test on calibration intercept and slope was 0.09.

### 2.2 Discrimination

We here calculate

+ The c-statistic: it is a rank order statistic for predictions against true outcomes. The concordance (c) statistic is the most commonly used performance measure to indicate the discriminative ability of generalized linear regression models. For a binary outcome, c is identical to the area under the Receiver Operating Characteristic (ROC) curve, which plots the sensitivity (true positive rate) against 1 – (false positive rate) for consecutive cutoffs for the probability of an outcome. Accurate predictions discriminate between those with and those without the outcome.

+ Discrimination slope: it can be used as a simple measure for how well subjects with and without the outcome are separated. It is calculated as the absolute difference in average predictions for those with and without the outcome. Visualization is readily possible with a box plot or a histogram, which will show less overlap between those with and those without the outcome for a better discriminating model. 
 

More details are in ["Assessing the performance of prediction models: a framework for some traditional and novel measures"](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3575184/) by Steyerberg et al. (2010);

+ Royston-Sauerbrei D statistic and R<sup>2</sup><sub>D</sub> as a measure of prognostic separation;  

  
#### 2.2.1 C-statistic 

<details>
  <summary>Click to expand code</summary>
```{r, discrimination,warning=FALSE,message=FALSE,echo=TRUE}

# Models


# useful objects
primary_event <- 1 # Set to 2 if cause 2 was of interest
horizon <- 5 # Set time horizon for prediction (here 5 years)

# C-index
# Development set (Apparent validation)

# Internal validation (bootstrapping)

# External validation
```
</details>

```{r, res_disc, fig.align='center',echo=FALSE}
alpha <- .05
k <- 2
res_discr_csh <- matrix(c(
  ## C-index
  # Development CSH1
  C_rdata,
  quantile(rboot$C_rboot, probs = alpha / 2),
  quantile(rboot$C_rboot, probs = 1 - alpha / 2),

  # Validation CSH1
  C_vdata,
  quantile(vboot$C_vboot, probs = alpha / 2),
  quantile(vboot$C_vboot, probs = 1 - alpha / 2),

  ## Time-dependent AUC
  # Development CSH1
  score_rdata$AUC$score$AUC,
  score_rdata$AUC$score$AUC -
    qnorm(1 - alpha / 2) * score_rdata$AUC$score$se,
  score_rdata$AUC$score$AUC +
    qnorm(1 - alpha / 2) * score_rdata$AUC$score$se,

  # Validation CSH1
  score_vdata$AUC$score$AUC,
  score_vdata$AUC$score$AUC -
    qnorm(1 - alpha / 2) * score_vdata$AUC$score$se,
  score_vdata$AUC$score$AUC +
    qnorm(1 - alpha / 2) * score_vdata$AUC$score$se
),
nrow = 2, ncol = 6, byrow = T,
dimnames = list(
  c("C-index", "Time dependent AUC"),
  rep(c("Estimate", "Lower .95", "Upper .95"), 2)
)
)
res_discr_csh <- round(res_discr_csh, k)
kable(res_discr_csh) %>%
  kable_styling("striped", position = "center") %>%
  add_header_above(c(
    " " = 1,
    "Development data" = 3,
    "Validation data" = 3
  ))
```

The time-dependent AUC at 5 years was 0.71 in the validation set.

#### 2.2.2 Plot Area under the curve(s) over the time
We plot the time-dependent AUCs over the follow-up time using development and validation data. 

<details>
  <summary>Click to expand code</summary>
```{r, plot_AUCs, warning=FALSE,echo=TRUE, eval=FALSE}

# Models --------------
fit_csh <- CSC(Hist(time, status_num) ~
age + size +
  ncat + hr_status,
data = rdata,
fitter = "cph"
)
primary_event <- 1 # Set to 2 if cause 2 was of interest

# AUCs development data
AUC_rdata <- Score(
  list("csh_development" = fit_csh),
  formula = Hist(time, status_num) ~ 1,
  cens.model = "km",
  data = rdata,
  conf.int = TRUE,
  times = seq(0.4, 5.1, 0.2),
  metrics = c("auc"),
  cause = primary_event
)

# AUCs validation data
AUC_vdata <- Score(
  list("csh_validation" = fit_csh),
  formula = Hist(time, status_num) ~ 1,
  cens.model = "km",
  data = vdata,
  conf.int = TRUE,
  times = seq(0.5, 5.01, 0.1),
  metrics = c("auc"),
  cause = primary_event
)

# Plot
par(las = 1, xaxs = "i", yaxs = "i")
oldpar <- par(mfrow = c(1, 2))
plot(AUC_rdata$times,
  AUC_rdata$AUC$score$AUC,
  type = "l",
  bty = "n",
  xlim = c(0, 5),
  ylim = c(.5, 1),
  lwd = 2,
  xlab = "Time (years)",
  ylab = "AUC",
  lty = 1
)
polygon(c(
  AUC_rdata$times,
  rev(AUC_rdata$times)
),
c(
  AUC_rdata$AUC$score$lower,
  rev(AUC_rdata$AUC$score$upper)
),
col = rgb(160, 160, 160, maxColorValue = 255, alpha = 100),
border = FALSE
)
lines(AUC_rdata$times,
  AUC_rdata$AUC$score$AUC,
  col = "black",
  lwd = 2,
  lty = 2
)
title("Development data", adj = 0)

# Validation data
plot(AUC_vdata$times,
  AUC_vdata$AUC$score$AUC,
  type = "l",
  bty = "n",
  xlim = c(0, 5),
  ylim = c(.5, 1),
  lwd = 2,
  xlab = "Time (years)",
  ylab = "AUC",
  lty = 1
)
polygon(c(
  AUC_vdata$times,
  rev(AUC_vdata$times)
),
c(
  AUC_vdata$AUC$score$lower,
  rev(AUC_vdata$AUC$score$upper)
),
col = rgb(160, 160, 160, maxColorValue = 255, alpha = 100),
border = FALSE
)
lines(AUC_vdata$times,
  AUC_vdata$AUC$score$AUC,
  col = "black",
  lwd = 2,
  lty = 2
)
title("Validation data", adj = 0)
```
</details>

```{r, plot_AUCs, warning=FALSE,echo=FALSE, eval=TRUE}
```

#### 2.2.3 Royston-Sauerbrei D statistic and R<sup>2</sup><sub>D</sub>

We calculate Royston-Sauerbrei D statistic and R<sup>2</sup><sub>D</sub> as a measure of explained variation

<details>
  <summary>Click to expand code</summary>
```{r, royston_R2D, fig.align='center', message=FALSE, warning=FALSE,echo=TRUE}
# function to calculate D and R2D
source(here::here("R/roystonD.R"))

# Models ----------
fit_csh <- CSC(Hist(time, status_num) ~
age + size +
  ncat + hr_status,
data = rdata,
fitter = "cph"
)


# useful objects
primary_event <- 1 # Set to 2 if cause 2 was of interest
horizon <- 5 # Set time horizon for prediction (here 5 years)

# Predicted risk estimation
rdata$pred <- predictRisk(fit_csh,
  cause = primary_event,
  times = horizon,
  newdata = rdata
)

vdata$pred <- predictRisk(fit_csh,
  cause = primary_event,
  times = horizon,
  newdata = vdata
)

# Rank based on predicted values at 5 years

rdata$rank <- rank(rdata$pred,
  na.last = TRUE,
  ties.method = "first"
)

vdata$rank <- rank(vdata$pred,
  na.last = TRUE,
  ties.method = "first"
)


res_rdata <- royston_R2D_cmprsk(
  pred = rdata$pred,
  data = rdata,
  primary_event = 1,
  time = rdata$time,
  status = rdata$status_num
)

res_vdata <- royston_R2D_cmprsk(
  pred = vdata$pred,
  data = vdata,
  primary_event = 1,
  time = vdata$time,
  status = vdata$status_num
)
```
</details>

```{r, res_D_R2D, fig.align='center',echo=FALSE}
k <- 2
res_D <- rbind(
  "Development data" = res_rdata,
  "Validation data" = res_vdata
)

res_D <- round(res_D, k)
kable(res_D) %>%
  kable_styling("striped", position = "center")
```

The R<sup>2</sup><sub>D</sub> showed how much of the observed variation in the outcome is explained by the prognostic model. The model explained 15% and 26% of the observed variation in the development and validation data, respectively.


### 2.3 Overall prediction error

We calculate the Brier Score, and the scaled Brier scale (also indicated by scaled Brier score or IPA) and the corresponding confidence intervals.

Some confidence intervals are calculated using the bootstrap percentile method.
```{r, bootstrap,warning=FALSE}
# Bootstrapping data
set.seed(20201214)
B <- 10 # number of bootstrap samples
rboot <- bootstraps(rdata, times = B)
vboot <- bootstraps(vdata, times = B)
```

```{r, overall, warning=FALSE}

# Models -------------------
fit_csh <- CSC(Hist(time, status_num) ~
age + size +
  ncat + hr_status,
data = rdata,
fitter = "cph"
)
fit_csc1 <- fit_csh$models$`Cause 1`
fit_csc2 <- fit_csh$models$`Cause 2`

# Overall performance measures ----------------
primary_event <- 1 # Set to 2 if cause 2 was of interest
horizon <- 5 # Set time horizon for prediction (here 5 years)

# Development data
score_rdata <- Score(
  list("csh_development" = fit_csh),
  formula = Hist(time, status_num) ~ 1,
  cens.model = "km",
  data = rdata,
  conf.int = TRUE,
  times = horizon,
  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

# Validation data
score_vdata <- Score(
  list("csh_validation" = fit_csh),
  formula = Hist(time, status_num) ~ 1,
  cens.model = "km",
  data = vdata,
  conf.int = TRUE,
  times = horizon,
  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

# Bootstrap ------
# Functions to expand data and calculate Brier, IPA and AUC in bootstrap
# samples.
# For Brier and AUC, bootstrap should be computationally faster when
# data has more than 2000 rows (see ?riskRegression::Score).
# Our data has 1000 row so we will need only bootstrap to calculate
# confidence intervals of the scaled Brier (IPA) since
# it is not provided by riskRegression::Score() function.


# Score functions in any bootstrap data
score_boot <- function(split) {
  Score(
    list("csh_validation" = fit_csh),
    formula = Hist(time, status_num) ~ 1,
    cens.model = "km",
    data = analysis(split),
    conf.int = TRUE,
    times = horizon,
    metrics = c("auc", "brier"),
    summary = c("ipa"),
    cause = primary_event,
    plots = "calibration"
  )
}

# Development data
rboot <- rboot %>% mutate(
  score = map(splits, score_boot),
  scaled_brier = map_dbl(score, function(x) {
    x$Brier$score[model == "csh_validation"]$IPA
  })
)
# Validation data
vboot <- vboot %>% mutate(
  score = map(splits, score_boot),
  scaled_brier = map_dbl(score, function(x) {
    x$Brier$score[model == "csh_validation"]$IPA
  })
)
```

```{r, res_ov, fig.align='center',echo=FALSE}
# Table overall measures
alpha <- .05
k <- 2 # number of digits
res_ov_csh <- matrix(unlist(c(
  # Brier score - apparent validation
  score_rdata$Brier$score[model == "csh_development"]$Brier,
  score_rdata$Brier$score[model == "csh_development"]$lower,
  score_rdata$Brier$score[model == "csh_development"]$upper,

  # Brier score - external validation
  score_vdata$Brier$score[model == "csh_validation"]$Brier,
  score_vdata$Brier$score[model == "csh_validation"]$lower,
  score_vdata$Brier$score[model == "csh_validation"]$upper,

  # IPA - apparent validation
  score_rdata$Brier$score[model == "csh_development"]$IPA,
  quantile(rboot$scaled_brier, probs = alpha / 2),
  quantile(rboot$scaled_brier, probs = 1 - alpha / 2),

  # IPA - external validation
  score_vdata$Brier$score[model == "csh_validation"]$IPA,
  quantile(vboot$scaled_brier, probs = alpha / 2),
  quantile(vboot$scaled_brier, probs = 1 - alpha / 2)
)),
nrow = 2, ncol = 6,
byrow = T,
dimnames =
  list(
    c("Brier", "scaled Brier"),
    rep(c("Estimate", "Lower .95 ", "Upper .95"), 2)
  )
)
res_ov <- round(res_ov_csh, 2) # Digits
kable(res_ov) %>%
  kable_styling("striped", position = "center") %>%
  add_header_above(c(
    " " = 1,
    "Development data" = 3,
    "Validation data" = 3
  ))
```

Note: unexpectedly, the point estimate for the Brier score is lower (thus better) and for the scaled Brier score is higher (thus better) in the validation data compared to the development data.

## Goal 3 -  Clinical utility

Clinical utility can be measured by the net benefit and plotted in a decision curve. Details about net benefit, decision curve calculation and interpretation are provided in the manuscript (see also the appendix) and its references.

<details>
  <summary>Click to expand code</summary>
```{r, dca, message=FALSE,warning=FALSE, fig.align='center',echo=TRUE,eval=FALSE}
# Run the stdca function to calculate the net benefit and the elements needed to develop decision curve analysis
source(here::here("R/stdca.R"))

# Models ------------------------------
fit_csh <- CSC(Hist(time, status_num) ~
age + size +
  ncat + hr_status,
data = rdata,
fitter = "cph"
)

# useful objects
primary_event <- 1 # Set to 2 if cause 2 was of interest
horizon <- 5 # Set time horizon for prediction (here 5 years)

# Development data
# calculation estimated risk
rdata$pred5 <- predictRisk(fit_csh,
  cause = primary_event,
  newdata = rdata,
  times = horizon
)
rdata <- as.data.frame(rdata)
dca_rdata <- stdca(
  data = rdata,
  outcome = "status_num",
  ttoutcome = "time",
  timepoint = horizon,
  predictors = "pred5",
  xstop = 0.35,
  ymin = -0.01,
  graph = FALSE,
  cmprsk = TRUE
)
# Decision curves plot
oldpar <- par(
  xaxs = "i",
  yaxs = "i",
  las = 1,
  mar = c(6.1, 5.8, 4.1, 2.1),
  mgp = c(4.25, 1, 0)
)
plot(dca_rdata$net.benefit$threshold,
  dca_rdata$net.benefit$pred5,
  type = "l",
  lwd = 2,
  lty = 1,
  xlab = "",
  ylab = "Net Benefit",
  xlim = c(0, 0.5),
  ylim = c(-0.10, 0.10),
  bty = "n",
  xaxt = "n"
)
legend("topright",
  c("Treat all", "Treat none", "Prediction model"),
  lwd = c(2, 2, 2),
  lty = c(1, 2, 1),
  col = c("darkgray", "black", "black"),
  bty = "n"
)
lines(dca_rdata$net.benefit$threshold,
  dca_rdata$net.benefit$none,
  type = "l",
  lwd = 2,
  lty = 4
)
lines(dca_rdata$net.benefit$threshold,
  dca_rdata$net.benefit$all,
  type = "l",
  lwd = 2,
  col = "darkgray"
)
axis(1,
  at = c(0, 0.1, 0.2, 0.3, 0.4, 0.5)
)
axis(1,
  pos = -0.145,
  at = c(0.1, 0.2, 0.3, 0.4, 0.5),
  labels = c("1:9", "1:4", "3:7", "2:3", "1:1")
)
mtext("Threshold probability", 1, line = 2)
mtext("Harm to benefit ratio", 1, line = 5)
title("Development data")
par(oldpar)


# Validation data
# Predicted probability calculation
vdata$pred5 <- predictRisk(fit_csh,
  cause = primary_event,
  newdata = vdata,
  times = horizon
)
vdata <- as.data.frame(vdata)
# Run decision curve analysis
# Development data
# Model without PGR
dca_vdata <- stdca(
  data = vdata,
  outcome = "status_num",
  ttoutcome = "time",
  timepoint = 5,
  predictors = "pred5",
  xstop = 0.45,
  ymin = -0.01,
  graph = FALSE,
  cmprsk = TRUE
)
# Decision curves plot
oldpar <- par(
  xaxs = "i",
  yaxs = "i",
  las = 1,
  mar = c(6.1, 5.8, 4.1, 2.1),
  mgp = c(4.25, 1, 0)
)
plot(dca_vdata$net.benefit$threshold,
  dca_vdata$net.benefit$pred5,
  type = "l",
  lwd = 2,
  lty = 1,
  xlab = "",
  ylab = "Net Benefit",
  xlim = c(0, 0.5),
  ylim = c(-0.10, 0.10),
  bty = "n",
  xaxt = "n"
)
lines(dca_vdata$net.benefit$threshold,
  dca_vdata$net.benefit$none,
  type = "l",
  lwd = 2,
  lty = 4
)
lines(dca_vdata$net.benefit$threshold,
  dca_vdata$net.benefit$all,
  type = "l",
  lwd = 2,
  col = "darkgray"
)
legend("topright",
  c("Treat all", "Treat none", "Prediction model"),
  lwd = c(2, 2, 2),
  lty = c(1, 2, 1),
  col = c("darkgray", "black", "black"),
  bty = "n"
)
axis(1,
  at = c(0, 0.1, 0.2, 0.3, 0.4, 0.5)
)
axis(1,
  pos = -0.145,
  at = c(0.1, 0.2, 0.3, 0.4, 0.5),
  labels = c("1:9", "1:4", "3:7", "2:3", "1:1")
)
mtext("Threshold probability", 1, line = 2)
mtext("Harm to benefit ratio", 1, line = 5)
title("Validation data")
par(oldpar)
```
</details>

```{r, dca, message=FALSE,warning=FALSE, fig.align='center',echo=FALSE,eval=TRUE}
```

If we choose a threshold of 20\%, the model had a net benefit of 0.011 in the development data. In the validation data, the model had a net benefit of 0.014 choosing a threshold of 20\%. 

## Reproducibility ticket

```{r repro_ticket, echo=TRUE}
sessionInfo()
```
